---
title: "F1 Dashboard"
author: "Michael L. Smith"
date: "`r format(Sys.time(), '%m.%d.%Y')`"
output: html_document
runtime: shiny
---

## Note to Marta and Jiuzhou: 
I cannot get the app to work by uploading to the `shinyapps.io`, however it does run on my local machine within this RMD file. I **think** I've uploaded the correct files that I'm using locally for it to run on your personal machine. If not please let me know! The knitted RMD file embeds the app, but the formatting gets weird and it's more difficult to navigate. If you run the `Shiny Formula 1 App!` section below the app **should** work (after loading packages of course).  

Link to code on github: (https://github.com/smit8854/Shiny_F1_app)

## Data Sources
1. [Ergast Api](http://ergast.com/mrd/)  
2. [Arkraieski Github](https://github.com/arkraieski/formula1data)
3. [Report by Yun Mai](https://rstudio-pubs-static.s3.amazonaws.com/276999_042092be8e31414f82ef4f41e31fe5c8.html)

## Load Libraries

```{r setup, include=TRUE, message=FALSE, warning=FALSE}

library(shiny)
library(tidyverse)
library(httr)
library(jsonlite)
library(gt)
library(lubridate)
library(gganimate)
library(gifski)
library(formula1data)
library(XML)
library(httr)
library(RCurl)
library(jsonlite)

library(devtools)
install_github("arkraieski/formula1data")


```


## Load Datasets
Using the `formula1data` package download the `Ergast API` data. These data include every F1 season from 1950 - 2021. For each season there are data for the race results, qualifying results, standings, driver information, constructor information, circuit information, lapt times, pit stops, and racing schedule.  

First we need to create a connection to the `Ergast API` using the `formula1data` package. This only needs to be done once! So after you run this you can comment it out.

```{r load data}
# createF1db()

```


## F1 Schedule Function
This is a modified version of the `getF1Schedule()` function from the `formula1data` package. Used to create the calendar tab in the shiny app. Run this code before you run shiny app.

```{r edit getF1schedule function}

getF1Schedule2 <- function(year){
  url <- paste0("https://ergast.com/api/f1/", year, ".json")
  sched <- fromJSON(content(GET(url), as = "text"))$MRData$RaceTable$Races
  sched <- sched %>%
    mutate_at(c("season", "round"), as.integer) %>%
    mutate(time = gsub("Z", "", .data$time),
           datetime = ymd_hms(paste(.data$date, .data$time)))
    # select(-.data$time)
  sched

}

```


## Shiny Formula 1 App!
This app has 4 main features:   
1. a tab for the season driver championship results  
2. a tab for the season constructor championship results  
3. a tab for the 2021 season calendar  
4. a tab for the 2021 race results so far  

I am not satisfied with how the app looks visually and the layout of a few features but it functions. Run the chunk of code below to view the shiny app in all it's glory. Note about the 2021 race results tab: it may look like it's doing nothing at first, but wait ~30 seconds and the gif should load with the race results table below!

```{r Shiny App, echo = FALSE}

ui <- navbarPage(title = "Formula 1 Dashboard",
                  tabPanel(title = "Season Driver Results",
                           numericInput(inputId = "season_driver", label = "Enter a year from 1950 - 2021", value = 2021, min = 1950, max = 2021),
                           plotOutput("season_driver"),
                           br(),
                           gt_output("driver_season"), br(),br()
                  ),
                  tabPanel(title = "Season Constructor  Results",
                           numericInput(inputId = "year", label = "Enter a year from 1958 - 2021", value = 2021, min = 1958, max = 2021),
                           plotOutput("constructor_plot"),
                           br(),
                           gt_output("season_table"), br(), br()
                  ),
                  # tabPanel(title = "Constructor History",
                  #         selectInput(inputId = "team", label = "Constructor Name", choices = constructor_results$Team),
                  #         gt_output("team_table")                          
                          # plotOutput("season_input")
                  # ),     
                  # tabPanel(title = "Driver History",
                  #            selectInput(inputId = "driver", label = "Driver Name", choices = drivers$surname),
                  #            tableOutput("driver_table")
                  # ),
                  tabPanel(title = "2021 Schedule",
                          gt_output(outputId = "round_table")
                  ),
                  tabPanel(title = "Race Results 2021",
                           # numericInput(inputId = "year_gif", label = "Enter a year from 1950 - 2021", value = 2020, min = 1950, max = 2021),
                           selectInput(inputId = "race_gif", label = "Select a round from the 2021 season", choices = c("Bahrain" = 1, "Imola" = 2),
                                       selected = NULL),
                           imageOutput("results_gif"),
                           br(),br(),br(),br(),br(),br(),
                           gt_output("results_table"),
                           br(), br()
                  )
)

server <- function(input, output) {
  
  season_driver <- reactive({
    getFinalF1Standings(input$season_driver)
  })
  
  season_team <- reactive({
    getFinalF1Standings(input$year, type = "constructor")
  })  
  
  schedule_21 <- getF1Schedule2(2021)
  
  schedule <- reactive({
    getF1Schedule(input$race_gif)
  })
  
  lapinfo <- reactive({
    getLapsByRace(2021, input$race_gif) %>% mutate(constructor = case_when(
      startsWith(driverId, "ham") ~ "Mercedes", startsWith(driverId, "bot") ~ "Mercedes",
      startsWith(driverId, "max") ~ "Red Bull", startsWith(driverId, "per") ~ "Red Bull",
      startsWith(driverId, "lec") ~ "Ferrari", startsWith(driverId, "sai") ~ "Ferrari",
      startsWith(driverId, "ric") ~ "McLaren", startsWith(driverId, "nor") ~ "McLaren",
      startsWith(driverId, "vet") ~ "Aston Martin", startsWith(driverId, "str") ~ "Aston Martin",
      startsWith(driverId, "gas") ~ "Alpha Tauri", startsWith(driverId, "tsu") ~ "Alpha Tauri",
      startsWith(driverId, "gio") ~ "Alfa Romeo", startsWith(driverId, "rai") ~ "Alfa Romeo",
      startsWith(driverId, "maz") ~ "Haas", startsWith(driverId, "mick") ~ "Haas",
      startsWith(driverId, "oco") ~ "Alpine", startsWith(driverId, "alo") ~ "Alpine",
      startsWith(driverId, "rus") ~ "Williams", startsWith(driverId, "lat") ~ "Williams"))
  })
  
team.color <- c("Mercedes" = "green",
                "Red Bull" = "blue",
                "Ferrari" = "red",
                "McLaren" = "orange",
                "Aston Martin" = "dark green",
                "Alpha Tauri" = "dark blue",
                "Alfa Romeo" = "dark red",
                "Haas" = "dark grey",
                "Alpine" = "cyan",
                "Williams" = "black")
  
  race_result <- reactive({
    getRaceResults(2021, input$race_gif)
  })
    
# Season Driver Tab
  observeEvent(input$season_driver, {
    
    # Plot
    output$season_driver <- renderPlot({
      # Fix overlap labels: some drivers are being dropped
      ggplot(season_driver()) +
        aes(x = position, y = points, label = driverId) +
        geom_text(check_overlap = F) +
        scale_x_continuous(breaks = c(1:99)) +
        labs(x = "Position", y = "Points", title = "Final Season Standings") +
        theme_bw() + theme(axis.text = element_text(size = 14), axis.title = element_text(size = 16))
    })
    # Table
    output$driver_season <- render_gt({
      season_driver() %>% 
        select(position, driverId, points, wins) %>% 
        gt() %>% 
        tab_style(locations = cells_body(columns = "driverId"), style = cell_text(transform = "capitalize")) %>% 
        tab_options(table.width = px(500)) %>%
        cols_align(align = "center", columns = TRUE) %>% 
        cols_label(position = "Position", points = "Points", wins = "Wins", driverId = "Driver")
    })
    
  })
  
# Season COnstructor Tab
  observeEvent(input$year,{
    
    # Plot
    output$constructor_plot <- renderPlot({
        ggplot(season_team()) +
          aes(x = position, y = points, label = constructorId) +
          geom_text(size = 5) +
          scale_x_continuous(breaks = c(1:30)) +
          labs(x = "Position", y = "Points", title = "Final Season Standings") +
          theme_bw() + theme(axis.text = element_text(size = 14), axis.title = element_text(size = 16))        
      })  
    
    # output$constructor_plot <- renderPlot({
    #   constructor_results %>% 
    #     filter(year == input$year) %>% 
    #     ggplot() +
    #       aes(x = Pos, y = Pts, label = Team) +
    #       geom_text() +
    #       scale_x_continuous(breaks = c(1:20)) +
    #       labs(title = "Final Season Standings")
    #   })
    
    # Table
    output$season_table <- render_gt({
      season_team() %>% 
        select(position, constructorId, points, wins) %>% 
        gt() %>% 
        tab_style(locations = cells_body(columns = "constructorId"), style = cell_text(transform = "capitalize")) %>% 
        tab_options(table.width = px(500)) %>%
        cols_align(align = "center", columns = TRUE) %>% 
        cols_label(position = "Position", points = "Points", wins = "Wins", constructorId = "Constructor")        
    })
    # Table using the other constructor results dataframe from the for loop
    # output$season_table <- render_gt({
    #   constructor_results %>% 
    #     filter(year == input$year) %>%
    #     gt() %>% 
    #     fmt_number(columns = "Pos", decimals = 0) %>% 
    #     cols_align(align = "center", columns = TRUE) %>%  
    #     cols_label(Pos = "Position", Pts = "Points", year = "Year")
    #   })
      
  })
  
# Constructor History Tab
  # observeEvent(input$team,{
  #  
  #   output$team_table <- render_gt({
  #     constructor_results %>% 
  #       filter(Team == input$team) %>%
  #       group_by(Pos) %>%
  #       summarize(num_finishes = n()) %>% 
  #       gt() %>% 
  #       fmt_number(columns = "Pos", decimals = 0) %>% 
  #       cols_align(align = "center", columns = TRUE) %>%         
  #       cols_label(Pos = "Position", num_finishes = "Number of Finishes")
  #     })
  # })
  # 
# Driver History Tab  
  output$driver_table <- render_gt({
    drivers
  })
  
# Calendar Tab
  output$round_table <- render_gt({
    schedule_21 %>% 
      mutate(circuit.name = Circuit$circuitName,
             City = Circuit$Location$locality,
             Country = Circuit$Location$country) %>% 
      select(round, raceName, circuit.name, City, Country, date, time) %>%     
      gt() %>%
      cols_align(align = "center", columns = TRUE) %>% 
      fmt_time(columns = vars(time), time_style = 4) %>% 
      fmt_date(columns = vars(date), date_style = 5) %>% 
      cols_label(round = "Round", raceName = "Race Name", circuit.name = "Circuit Name", date = "Date", time = "Local Race Time")
    })

# Animation Tab
  output$results_gif <- renderImage({
    outfile <- tempfile(fileext = ".gif")
    
    gif <- ggplot(lapinfo()) +
      aes(x = lap, y = position, label = driverId, color = constructor) +
      geom_text(size = 5) +
      scale_y_reverse(breaks = c(1:30)) +
      labs(title = "Lap {frame_time}") +
      scale_color_manual(values = team.color) +
      theme_bw() + theme(axis.text = element_text(size = 14), axis.title = element_text(size = 16)) +
      transition_time(lap) +
      ease_aes("linear")
    
    anim_save("outfile.gif", animate(gif, nframes = 300, fps = 15, end_pause = 100))
    
    list(src = "outfile.gif",
         contentType = 'image/gif'
         # width = 400,
         # height = 300,
         # alt = "This is alternate text"
    )},deleteFile = TRUE) 
  
  # result table
  output$results_table <- render_gt({
    race_result() %>% 
      mutate(driver.familyname = Driver$familyName,
             constructor.name = Constructor$name,
             time = Time$time) %>% 
      select(positionText, points, driver.familyname, constructor.name, time, grid) %>% 
      gt() %>% 
      cols_align(align = "center", columns = TRUE) %>% 
      cols_label(positionText = "Position", points = "Points", driver.familyname = "Driver", constructor.name = "Constructor",
                 time = "Total Race Time", grid = "Qualifying Position")
      
    })
  
}



shinyApp(server = server, ui = ui)

```


## Conclusions
The shiny app I made needs a lot of work but it is functional in some aspects. I would like to make it prettier using HTML + CSS code, but I was struggling to implement these aesthetics with the rest of my shiny app code. There is also some tweaking I would want to do for the figures and the how the figures/tables are laid out in a tab.



<!-- ########## JUNK ################## -->

```{r TESTING PLOTS & TABLES, include = FALSE, eval = FALSE}

################## JUNK ##########################
# Test different plots, tables, etc., to possibly implement into shiny app.

season <- 2010
final.2010 <- getFinalF1Standings(season, type = "driver") 

ggplot(final.2010) + 
  aes(x = points, y = position, label = driverId) +
  geom_text() +
  scale_y_reverse()

getFinalF1Standings(season, type = "constructor") 
ggplot(final.2010) + 
  aes(x = points, y = position, label = constructorId) +
  geom_text() +
  scale_y_reverse()

constructor_results %>% 
  filter(year == 1985) %>% 
  ggplot() +
    aes(x = Pos, y = Pts, color = Team) +
    geom_point(size = 2) +
    scale_x_continuous(breaks = c(1:20))


constructor_results %>% 
  filter(year == 1985) %>%
  gt() %>% 
  fmt_number(columns = "Pos", decimals = 0)

constructor_results %>% 
  filter(Team == "Ferrari") %>%
  arrange(Pos) %>% 
  gt() %>% 
  fmt_number(columns = "Pos", decimals = 0)

constructor_results %>% 
  filter(Team == "Ferrari") %>%
  group_by(Pos) %>%
  summarize(num_finishes = n())



# drivers_merged <- left_join(driver_standings, drivers, by = "driverId")
# 
# drivers_merged %>%
#   filter(forename == "Lewis" & surname == "Hamilton") %>%
#   group_by(driverRef, code, position) %>%
#   count()


calendar_2021 <- getF1Schedule2(2021)

calendar_2021 %>% 
  mutate(circuit.name = Circuit$circuitName,
         City = Circuit$Location$locality,
         Country = Circuit$Location$country) %>% 
  select(round, raceName, circuit.name, City, Country, date, time) %>%
  gt() %>% 
  fmt_time(columns = vars(time), time_style = 4) %>% 
  fmt_date(columns = vars(date),
               date_style = 5) %>% 
  cols_align(align = "center", columns = TRUE) %>% 
  cols_label(round = "Round", raceName = "Grand Prix", circuit.name = "Circuit Name", date = "Date", time = "Local Race Time")

test <- calendar_2021$Circuit

calendar_2021$Circuit$Location$locality
calendar_2021$Circuit$Location$country


########### Animated gif
# create a gif that shows a race result over each lap

race2.2020 <- getLapsByRace(2020, 2)
race7.2012 <- getLapsByRace(2012, 7)

# driver.teams.21 <- list("Mercedes" = c("hamilton", "bottas"),
#                         "Red Bull" = c("max_verstappen", "albon"),
#                         "Ferrari" = c("leclerc", "vettel"),
#                         "McLaren" = c("norris", "sainz"),
#                         "Racing Point" = c("perez", "stroll"),
#                         "Alpha Tauri" = c("gasly", "kvyat"),
#                         "Alfa Romeo" = c("giovinazzi", "raikkonen"),
#                         "Haas" = c("grosjean", "kevin_magnussen"),
#                         "Renalut" = c("ocon", "ricciardo"),
#                         "Williams" = c("russell", "latifi"))
# 
# driver.teams.21 <- list("hamilton" = "Mercedes",
#                         "bottas" = "Mercedes",
#                         "max_verstappen" = "Red Bull",
#                         "albon" = "Red Bull",
#                         "leclerc" = "Ferrari",
#                         "vettel" = "Ferrari",
#                         "norris" = "McLaren",
#                         "sainz" = "McLarne",
#                         "perez" = "Racing Point",
#                         "stroll" = "Racing Point",
#                         "gasly" = "Alpha Tauri",
#                         "kvyat" = "Alpha Tauri",
#                         "giovinazzi" = "Alfa Romeo",
#                         "raikkonen" = "Alfa Romeo",
#                         "grosjean" = "Haas",
#                         "kevin_magnussen" = "Haas",
#                         "ocon" = "Renault",
#                         "ricciardo" = "Renault", 
#                         "russell" = "Williams",
#                         "latifi" = "Williams")

race2.2020 <- race2.2020 %>% 
  mutate(constructor = case_when(
    startsWith(driverId, "ham") ~ "Mercedes", startsWith(driverId, "bot") ~ "Mercedes",
    startsWith(driverId, "max") ~ "Red Bull", startsWith(driverId, "alb") ~ "Red Bull",
    startsWith(driverId, "lec") ~ "Ferrari", startsWith(driverId, "vet") ~ "Ferrari",
    startsWith(driverId, "sai") ~ "McLaren", startsWith(driverId, "nor") ~ "McLaren",
    startsWith(driverId, "per") ~ "Racing Point", startsWith(driverId, "str") ~ "Racing Point",
    startsWith(driverId, "gas") ~ "Alpha Tauri", startsWith(driverId, "kvy") ~ "Alpha Tauri",
    startsWith(driverId, "gio") ~ "Alfa Romeo", startsWith(driverId, "rai") ~ "Alfa Romeo",
    startsWith(driverId, "kev") ~ "Haas", startsWith(driverId, "gro") ~ "Haas",
    startsWith(driverId, "oco") ~ "Renault", startsWith(driverId, "ric") ~ "Renault",
    startsWith(driverId, "rus") ~ "Williams", startsWith(driverId, "lat") ~ "Williams"))

team.color <- c("Mercedes" = "dark green",
                "Red Bull" = "blue",
                "Ferrari" = "red",
                "McLaren" = "orange",
                "Racing Point" = "pink",
                "Alpha Tauri" = "dark blue",
                "Alfa Romeo" = "dark red",
                "Haas" = "grey",
                "Renault" = "yellow",
                "Williams" = "black")

race.gif <-ggplot(race2.2020) +
      aes(x = lap, y = position, label = driverId, color = constructor) +
      geom_text(size = 5) +
      scale_y_reverse(breaks = c(1:40)) +
      labs(title = "Styrian Grand Prix 2020", subtitle = "Lap: {frame_time}") +
      scale_color_manual(values = team.color) +
      theme_bw() +
      transition_time(lap) +
      ease_aes("linear")
      
animate(race.gif, renderer = gifski_renderer())


canada2012 <- 
  ggplot(race7.2012) +
  aes(x = lap, y = position, label = driverId) +
    geom_text() +
    scale_y_reverse(breaks = c(1:30)) +
    labs(title = "Canadian Grand Prix 2012", subtitle = "Lap {frame_time} of 71") +
    transition_states(states = as.factor(position), transition_length = 25) +
    transition_time(lap) +
    ease_aes("linear")
      
animate(canada2012, renderer = gifski_renderer())

dummy.results <- getRaceResults(2021,2)



```



